---
title: "Climate_Change"
author: "Slavoljub Petkovic"
date: "May 21, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

This project is meant to shed light on several questions related to climate changes. I will be using set of data from Kaggle.com which contains four files with temperature details in period of time between 1850 and 2013:

1. Global Temperatures (GlobalTemperatures.csv)
2. Global Average Land Temperature by Country (GlobalLandTemperaturesByCountry.csv)
3. Global Average Land Temperature by State (GlobalLandTemperaturesByState.csv)
4. Global Land Temperatures by Major City (GlobalLandTemperaturesByMajorCity.csv)

[Data can be located here] (https://www.kaggle.com/berkeleyearth/climate-change-earth-surface-temperature-data)

In this research, I will attempt to answer few questions below:

1. Does the rate of change over the continent correlate with changes over the states around the world?
2. What are the top 5 hottest and top 5 coldest major cities in the world?
3. What kind of climate we can expect in next 5 years, 10 years, 50 years and 100 years based on predictive analysis?
4. Is there an active global warming?

I will be providing the analysis that might include dynamic graphs and charts with short explanation of what each of them means and why is it important for us.

## Data Wrangling

* Removing irrelevant columns and rows
* Standardizing Column Names
* Exploring the data for missing values and NA
* Formatting values for consistency
* Explore outliers with min and max to see the extremes 
* Merging data to streamline the analysis flow

```{r ilibraries}
library(lubridate)
library(magrittr)
library(dplyr)
library(tidyr)
library(readr)
```

#### Global Temperatures - (GlobalTemperatures.csv)
This data set contains next columns: dt, LandAverageTemperature, LandAverageTemperatureUncertainty, LandMaxTemperature, LandMaxTemperatureUncertainty, LandMinTemperature, LandMinTemperatureUncertainty, LandAndOceanAverageTemperature, LandAndOceanAverageTemperatureUncertainty

* Removing data prior 1850 to be removed. 

```{r TempGlobal, message=TRUE}
# Loading data set
GlobalTemperatures <- read_csv("C:/GIT/R/Climate_Change/Data_Source/GlobalTemperatures.csv")
View(GlobalTemperatures)
# Filtering data range
GlobalTemperatures_Mod <- filter(GlobalTemperatures, dt>= "1849-12-31")                      
# Saving edited file
write.table(GlobalTemperatures_Mod, "Global_Temp.csv")                                       
# Removing original data frame
rm(GlobalTemperatures)                                                                       
```

#### Global Average Land Temperature by Country (GlobalLandTemperaturesByCountry.csv)
This set contains next columns: dt, AverageTemperature, AverageTemperautreUncertanity and Country.

* Removing data prior 1850 to be removed. 

```{r TempByCountry}
# Loading data set
GlobalLandTemperaturesByCountry <- read_csv("C:/GIT/R/Climate_Change/Data_Source/GlobalLandTemperaturesByCountry.csv")
View(GlobalLandTemperaturesByCountry)
# Filtering data range
GlobalLandTemperaturesByCountry_Mod <- filter(GlobalLandTemperaturesByCountry, dt>= "1849-12-31")
# Saving edited file
write.table(GlobalLandTemperaturesByCountry_Mod, "Global_Temp_Country.csv")                                       
# Removing original data frame
rm(GlobalLandTemperaturesByCountry)    
```

#### Global Average Land Temperature by State (GlobalLandTemperaturesByState.csv)
This set contains next columns: dt, AverageTemperature, AverageTemperatureUncertainty, State,	Country.

* Removing data prior 1850 to be removed. 

```{r TempByStates}
# Loading data set
GlobalLandTemperaturesByState <- read_csv("C:/GIT/R/Climate_Change/Data_Source/GlobalLandTemperaturesByState.csv")
View(GlobalLandTemperaturesByState)
# Filtering data range
GlobalLandTemperaturesByState_Mod <- filter(GlobalLandTemperaturesByState, dt>= "1849-12-31")
# Saving edited file
write.table(GlobalLandTemperaturesByState_Mod, "Global_Temp_State.csv")                                       
# Removing original data frame
rm(GlobalLandTemperaturesByState)
```

#### Global Land Temperatures by Major City (GlobalLandTemperaturesByMajorCity.csv)
This set contains next columns: dt, AverageTemperature, AverageTemperatureUncertainty, City, Country, Latitude, Longitude.

* Removing data prior 1850 to be removed. 

```{r TempByMajorCity}
# Loading data set
GlobalLandTemperaturesByMajorCity <- read_csv("C:/GIT/R/Climate_Change/Data_Source/GlobalLandTemperaturesByMajorCity.csv")
View(GlobalLandTemperaturesByMajorCity)
# Filtering data range
GlobalLandTemperaturesByMajorCity_Mod <- filter(GlobalLandTemperaturesByMajorCity, dt>= "1849-12-31")
# Saving edited file
write.table(GlobalLandTemperaturesByMajorCity_Mod, "GlobalLaandTempBy_MCity.csv")                                       
# Removing original data frame
rm(GlobalLandTemperaturesByMajorCity)
```
