---
title: "Climate_Change"
author: "Slavoljub Petkovic"
date: "May 21, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

This project is meant to shed light on several questions related to climate changes. I will be using set of data from Kaggle.com which contains four files with temperature details in period of time between 1850 and 2013:

1. Global Temperatures (GlobalTemperatures.csv)
2. Global Average Land Temperature by Country (GlobalLandTemperaturesByCountry.csv)
3. Global Average Land Temperature by State (GlobalLandTemperaturesByState.csv)
4. Global Land Temperatures by Major City (GlobalLandTemperaturesByMajorCity.csv)

[Data can be located here] (https://www.kaggle.com/berkeleyearth/climate-change-earth-surface-temperature-data)

In this research, I will attempt to answer few questions below:

1. Does the rate of change over the continent correlate with changes over the states around the world?
2. What are the top 5 hottest and top 5 coldest major cities in the world?
3. What kind of climate we can expect in next 5 years, 10 years, 50 years and 100 years based on predictive analysis?
4. Is there an active global warming?

I will be providing the analysis that might include dynamic graphs and charts with short explanation of what each of them means and why is it important for us.

## Data Wrangling

* Removing irrelevant columns and rows
* Standardizing Column Names
* Exploring the data for missing values and NA
* Formatting values for consistency
* Explore outliers with min and max to see the extremes 
* Merging data to streamline the analysis flow

#### Global Temperatures - (GlobalTemperatures.csv)
This data set contains next columns: dt, LandAverageTemperature, LandAverageTemperatureUncertainty, LandMaxTemperature, LandMaxTemperatureUncertainty, LandMinTemperature, LandMinTemperatureUncertainty, LandAndOceanAverageTemperature, LandAndOceanAverageTemperatureUncertainty

* Data prior 1850 to be removed. 
* dt - dates to be standardized.

```{r TempGlobal}

```

#### Global Average Land Temperature by Country (GlobalLandTemperaturesByCountry.csv)
This set contains next columns: dt, AverageTemperature, AverageTemperautreUncertanity and Country.

* Data prior 1850 to be removed.
* dt - dates to be standardized. 
* AverageTemperature and AverageTemperautreUncertanity - missing values to be populated with average values for country for which values are missing.  

```{r TempByCountry}

```

#### Global Average Land Temperature by State (GlobalLandTemperaturesByState.csv)
This set contains next columns: dt, AverageTemperature, AverageTemperatureUncertainty, State,	Country.
* dt - dates to be standardized.
* AverageTemperature and AverageTemperatureUncertainty missing the values for some of the year a and will be populated with average values for state for which values are missing.

```{r TempByStates}

```

#### Global Land Temperatures by Major City (GlobalLandTemperaturesByMajorCity.csv)
This set contains next columns: dt, AverageTemperature, AverageTemperatureUncertainty, City, Country, Latitude, Longitude.

* Data prior 1850 to be removed.
* dt - dates to be standardized. 
* AverageTemperature and AverageTemperautreUncertanity - missing values to be populated with average values for city for which values are     missing.

```{r TempByCity}

```

#### Combining the data in to one data set
